---
title: 'Customer Churn'
author: "Samer Alzaim"
date: " "
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    theme: lumen
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
editor_options: 
  chunk_output_type: inline
---

```{=html}

<style type="text/css">

/* Cascading Style Sheets (CSS) is a stylesheet language used to describe the presentation of a document written in HTML or XML. it is a simple mechanism for adding style (e.g., fonts, colors, spacing) to Web documents. */

h1.title {  /* Title - font specifications of the report title */
  font-size: 24px;
  font-weight: bold;
  color: navy;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}
h4.author { /* Header 4 - font specifications for authors  */
  font-size: 18px;
  font-family: system-ui;
  color: navy;
  text-align: center;
  font-weight: bold;
}
h4.date { /* Header 4 - font specifications for the date  */
  font-size: 18px;
  font-family: system-ui;
  color: DarkBlue;
  text-align: center;
  font-weight: bold;
}
h1 { /* Header 1 - font specifications for level 1 section title  */
    font-size: 22px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: center;
    font-weight: bold;
}
h2 { /* Header 2 - font specifications for level 2 section title */
    font-size: 20px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
    font-weight: bold;
}

h3 { /* Header 3 - font specifications of level 3 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - font specifications of level 4 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

</style>
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("tidyverse")) {
   install.packages("tidyverse")
library(tidyverse)
}
if (!require("GGally")) {
   install.packages("GGally")
library(GGally)
}
if (!require("glmnet")) {
   install.packages("glmnet")
library(glmnet)
}
if (!require("caret")) {
   install.packages("caret")
library(caret)
}
if (!require("MASS")) {
   install.packages("MASS")
library(MASS)
}
if (!require("mlbench")) {
   install.packages("mlbench")
library(mlbench)
}
if (!require("pROC")) {
   install.packages("pROC")
library(pROC)
}
if (!require("plotly")) {
   install.packages("plotly")
library(plotly)
}
if (!require("pander")) {
   install.packages("pander")
library(pander)
}
if (!require("e1071")) {
   install.packages("e1071")
library(e1071)
}
### 
knitr::opts_chunk$set(echo = TRUE,       
                      warning = FALSE,    
                      results = TRUE,    
                      message = FALSE,
                      comment = NA
                      )  
```


```{r}
data1 <- read.table("https://pengdsci.github.io/datasets/ChurnData/Customer-Chrn-dataset.txt", 
                   header = TRUE, sep = ",", stringsAsFactors = FALSE)
```

# Background

In this step, we will look at numeric variable and fit the model to predict this variable. the dat set has 2 numeric variables, Monthly_Charges and Total_Charges. We will use Monthy_Scharges as our res-ponder variable.

# Fitting the Model

from the graph below, it seems that we only have 2 variables that contribute to the prediction of Monthly_Charges.

``` {r}

# Set seed for reproducibility
set.seed(654321)  

# Prepare features (X) and target (y)
X <- as.matrix(data1[, -12])  # Features (all columns except the target) 
y <- data1$Monthly_Charges  # Ensure numeric

# Split the data into training and testing sets
train_index <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[train_index, ]
X_test <- X[-train_index, ]
y_train <- y[train_index]
y_test <- y[-train_index]

# Standardize the data (important for regularization)
preprocess_params <- preProcess(X_train, method = c("center", "scale"))
X_train <- predict(preprocess_params, X_train)
X_test <- predict(preprocess_params, X_test)

# Fit Linear Regression Models with Regularization
fit_lasso <- glmnet(X_train, y_train, alpha = 1)       # Lasso regression  
fit_ridge <- glmnet(X_train, y_train, alpha = 0)       # Ridge regression  
fit_elastic_net <- glmnet(X_train, y_train, alpha = 0.5) # Elastic net  

# Cross-validation
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1)   
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0)   
cv_elastic_net <- cv.glmnet(X_train, y_train, alpha = 0.5)  

```


```{r}

par(mar=c(5,4,6,3))
# Plot coefficient path
plot(fit_lasso, xvar = "lambda", label = TRUE,
     lwd = 1.5,
     main = "Coefficient Path Analysis: LASSO",
     cex.main = 0.9,
     col = rainbow(10))
abline(v = 1, col = "purple", lty = 4, lwd = 2)
abline(v = -1, col = "steelblue", lty = 2, lwd = 2)

```

##RMSE Plot
Looking a RMSE plot we can see that best lambda is between 1 and zero, consideing lower lmbda will not further improve the RMSE.

```{r}
par(mar=c(5,4,6,3))
##
plot(cv_lasso, main = "RMSE Plot: LASSO",
     cex.main = 0.9)
```

## Optimal Lambda

Lasso and Elastic net have lower lambda comapred to Ridge

```{r}

# Cross-validation to find the best lambda
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1)
cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0)
cv_elastic_net <- cv.glmnet(X_train, y_train, alpha = 0.5)
##
# Extract coefficients for the best lambda
best.lasso.lambda <- cv_lasso$lambda.min
best.ridge.lambda <- cv_ridge$lambda.min
best.elastic.net.lambda <- cv_elastic_net$lambda.min
##
# Lasso Regression (L1 Regularization): 
# CAUTION: model formula differs from the regular regression formula 
lasso_model.opt <- glmnet(X_train, y_train, alpha = 1,      # lasso regression 
                      lambda = best.lasso.lambda)   # useser selected alpha, optimal lambda
                                      # can be obtained through CV (see below)
lasso_predictions.opt <- predict(lasso_model.opt, s = best.lasso.lambda, # user selected lambda value 
                                                                          # (regularization paremeter)
                             newx = X_test)  # test data set 
# The following RMSE of prediction serves as a validation - one step validation
lasso_rmse.opt <- sqrt(mean((y_test - lasso_predictions.opt)^2))   
                                          
# Ridge Regression (L2 Regularization)
ridge_model.opt <- glmnet(X_train, y_train, alpha = 0, lambda = best.ridge.lambda)
ridge_predictions.opt <- predict(ridge_model.opt, s = best.ridge.lambda, newx = X_test)
ridge_rmse.opt <- sqrt(mean((y_test - ridge_predictions.opt)^2))

# Elastic Net (Combination of L1 and L2)
elastic_net_model.opt <- glmnet(X_train, y_train, alpha = 0.5, lambda = 0.1)
elastic_net_predictions.opt <- predict(elastic_net_model.opt, s = 0.1, newx = X_test)
elastic_net_rmse.opt <- sqrt(mean((y_test - elastic_net_predictions.opt)^2))

RMSE.opt = cbind(LASSO.opt = lasso_rmse.opt, 
                 Ridge.opt =  ridge_rmse.opt, 
                 Elasticnet.opt = elastic_net_rmse.opt)
pander(RMSE.opt)

```

## Lasso Formula

```{r}
##lasso
# Extract coefficients for the best lambda
best_lambda.lasso <- cv_lasso$lambda.min
coefficients.lasso <- coef(cv_lasso, s = best_lambda.lasso)
# Reconstruct the model equation
intercept.lasso <- coefficients.lasso[1]
betas.lasso <- coefficients.lasso[-1]
cat("Model equation: y =", round(intercept.lasso,4), "+", 
paste(round(betas.lasso,4), colnames(X), sep = "*", collapse = " + "), "\n")

```

## Ridge Formula

### FUll model formula.
```{r}

##ridge

# Extract coefficients for the best lambda
best_lambda.ridge <- cv_ridge$lambda.min
coefficients.ridge <- coef(cv_ridge, s = best_lambda.ridge)
# Reconstruct the model equation
intercept.ridge <- coefficients.ridge[1]
betas.ridge <- coefficients.ridge[-1]
cat("Model equation: y =", round(intercept.ridge,4), "+", 
paste(round(betas.ridge,4), colnames(X), sep = "*", collapse = " + "), "\n")

```

### Short Model Formula with only variables with Coeficnet value greater than zero

```{r}

best_lambda.net <- cv_ridge$lambda.min

# Extract coefficients at best lambda
coefficients.net <- coef(cv_ridge, s = best_lambda.net)

# Convert to a numeric vector
coeff_values <- as.numeric(coefficients.net)
coeff_names <- rownames(coefficients.net)

# Remove terms where coefficients are zero
nonzero_indices <- which(coeff_values != 0)
nonzero_coeffs <- coeff_values[nonzero_indices]
nonzero_names <- coeff_names[nonzero_indices]

# Construct model equation
intercept.net <- round(nonzero_coeffs[1], 4)  # Intercept
betas.net <- round(nonzero_coeffs[-1], 4)  # Coefficients
predictors.net <- nonzero_names[-1]  # Predictor names

# Display final equation
cat("Model equation: y =", intercept.net, 
    ifelse(length(betas.net) > 0, paste("+", paste(betas.net, "*", predictors.net, collapse = " + ")), ""), "\n")
    
```

## Elastic Net Formula

### FUll model formula.
```{r}

##ElasticNet
# Extract coefficients for the best lambda
best_lambda.net <- cv_elastic_net$lambda.min
coefficients.net <- coef(cv_elastic_net, s = best_lambda.net)
# Reconstruct the model equation
intercept.net <- coefficients.net[1]
betas.net<- coefficients.net[-1]
cat("Model equation: y =", round(intercept.net,4), "+", 
paste(round(betas.net,4), colnames(X), sep = "*", collapse = " + "), "\n")

```


### Short Model Formula with only variables with Coeficnet value greater than zero

```{r}

best_lambda.net <- cv_elastic_net$lambda.min

# Extract coefficients at best lambda
coefficients.net <- coef(cv_elastic_net, s = best_lambda.net)

# Convert to a numeric vector
coeff_values <- as.numeric(coefficients.net)
coeff_names <- rownames(coefficients.net)

# Remove terms where coefficients are zero
nonzero_indices <- which(coeff_values != 0)
nonzero_coeffs <- coeff_values[nonzero_indices]
nonzero_names <- coeff_names[nonzero_indices]

# Construct model equation
intercept.net <- round(nonzero_coeffs[1], 4)  # Intercept
betas.net <- round(nonzero_coeffs[-1], 4)  # Coefficients
predictors.net <- nonzero_names[-1]  # Predictor names

# Display final equation
cat("Model equation: y =", intercept.net, 
    ifelse(length(betas.net) > 0, paste("+", paste(betas.net, "*", predictors.net, collapse = " + ")), ""), "\n")
    
```



```{r}


# Prepare data
X <- as.matrix(data1[, c("Term", "Total_Charges")])  # Predictors
y <- data1$Monthly_Charges  # Response

# Create interaction term
interaction_term <- X[, "Term"] * X[, "Total_Charges"]
X_with_interactions <- cbind(X, interaction_term)

# Fit glmnet model
fit <- glmnet(X_with_interactions, y, alpha = 1)  # this fits LASSO regression
# Extracts coefficients at lambda = 0.1
# 
#coef(fit, s = 0.1)    
# Cross-validation
cv_fit <- cv.glmnet(X_with_interactions, y, alpha = 1)

# Plot the cross-validation results for manually select the value of lambda
#plot(cv_fit)

# Best lambda value: automatically selected optimal lambda
best_lambda <- cv_fit$lambda.min
#print(best_lambda)

# Refit the model with the best lambda
final_model <- glmnet(X_with_interactions, y, alpha = 1, lambda = best_lambda)
coef(final_model, s = best_lambda)

```

# Support Vector Regression


```{r}

#####
# Split data into features (X) and target (y)
X <- data1[, -12]  # All columns except the target variable
y <- data1[, 12]   # Target variable (median house value)
#####
# Split data into training and testing sets
set.seed(123)
train.index <- sample(1:nrow(data1), 0.8 * nrow(data1))
X.train <- X[train.index, ]
y.train <- y[train.index]
X.test <- X[-train.index, ]
y.test <- y[-train.index]
#####
data1.train <- data1[train.index, ]
data1.test <- data1[-train.index, ]
#####
## Set up custom cross-validation control
tune.control <- tune.control(
  cross = 5,  # Use 5-fold cross-validation, the default is 10-fold cross-validation
  nrepeat = 1 # Number of repetitions (for repeated cross-validation)
)
#####

```

## Check number of values and colums and the nature of the table
800 observations with 13 columns which matches our design. also, data type is data.frame which what we are looking for

```{r}

print(dim(X.train))   # Should match the number of rows and columns expected
print(length(y.train))  # Should match the number of rows in X.train
print(class(X.train))  # Should be "matrix" or "data.frame"
print(class(y.train))  # Should be "numeric" or "integer"

print(class(X.train))  # Should be "data.frame" or "matrix"
print(class(y.train))  # Should be "numeric" or "integer"

print(sapply(X.train, class))  # Check each column's data type

```

## check if any colum have value that is not integer or numeric
there are many columns that have character values which can not be processed by SVM

```{r}

# Identify character columns
char_cols <- sapply(X.train, is.character)
print(names(X.train)[char_cols])  # Show character column names

# Convert character columns to factors, then to numeric
X.train[char_cols] <- lapply(X.train[char_cols], function(col) as.numeric(as.factor(col)))
X.test[char_cols] <- lapply(X.test[char_cols], function(col) as.numeric(as.factor(col)))

```

## Check if all charecter values converted to numeric
all values are numeric and integers.

```{r}

print(sapply(X.train, class))  # All should be numeric or integer

```

## Checking for missing values
No missing values

```{r}

# Check for missing values (NA) in X.train and y.train
print(sum(is.na(X.train)))  # Should be 0
print(sum(is.na(y.train)))  # Should be 0

# Check for infinite values
print(sum(is.infinite(as.matrix(X.train))))  # Should be 0
print(sum(is.infinite(y.train)))  # Should be 0

```
## RUN SVM manually first
It run without errors

```{r}

test.svm <- svm(X.train, y.train, 
                type = "eps-regression",  
                kernel = "radial", 
                cost = 1, gamma = 0.01, epsilon = 0.1)

print(test.svm)

```

## Now we run "tune" with small grid to see if it is woriking.
from the below, it seems it is working.

```{r}

tune.RBF <- tune(svm, train.x = X.train, train.y = y.train, 
                 ranges = list(epsilon = c(0.1, 0.2), 
                               cost = c(1, 10), 
                               gamma = c(0.01, 0.1)), 
                 tunecontrol = tune.control(sampling = "cross", cross = 3))  

print(tune.RBF$best.parameters)

```

```{r}

# Perform grid search for hyperparameter tuning: RBF kernel is used
tune.RBF <- tune(svm, train.x = X.train, train.y = y.train, 
                    ranges = list(epsilon = seq(0.1, 0.5, 0.1), 
                                  cost = c(1, 10, 100), 
                                  gamma = c(0.01, 0.1, 1)), # Hyperpar in RBF
                    tunecontrol = tune.control(sampling = "cross", 
                                               cross = 5) # 5-fold cross-validation
                    )
####
# Display the best parameters
#print(tune.result$best.parameters)
#####
# Train the final model using the best parameters
final.RBF<- svm(X.train, y.train, 
                   type = "eps-regression",  # Use "nu-regression" for nu-SVR
                   kernel = "radial", 
                   epsilon = tune.RBF$best.parameters$epsilon, 
                   cost = tune.RBF$best.parameters$cost, 
                   gamma = tune.RBF$best.parameters$gamma)
#####
# Make predictions on the test set
pred.RBF <- predict(final.RBF, X.test)

# Evaluate performance
mse.RBF <- mean((y.test - pred.RBF)^2)    # mean square error
mae.RBF <- mean(abs(y.test - pred.RBF))   # mean absolute error

#### linear kernel

# Perform grid search for hyperparameter tuning: RBF kernel is used
tune.lin <- tune(svm, train.x = X.train, train.y = y.train, 
                    ranges = list(epsilon = seq(0.1, 0.5, 0.1), 
                                  cost = c(1, 10, 100)), 
                    tunecontrol = tune.control(sampling = "cross", 
                                               cross = 5) # 5-fold cross-validation
                    )
####
# Display the best parameters
#print(tune.result$best.parameters)
#####
# Train the final model using the best parameters
final.lin<- svm(X.train, y.train, 
                   type = "eps-regression",  # Use "nu-regression" for nu-SVR
                   kernel = "radial", 
                   epsilon = tune.lin$best.parameters$epsilon, 
                   cost = tune.lin$best.parameters$cost)
#####
# Make predictions on the test set
pred.lin <- predict(final.lin, X.test)

# Evaluate performance
mse.lin <- mean((y.test - pred.lin)^2)    # mean square error
mae.lin <- mean(abs(y.test - pred.lin))   # mean absolute error

```

```{r}

# Convert relevant columns in both train and test to factors
factor_columns <- c('Marital_Status', 'Multiple_line', 'Internet_service', 
                    'Technical_support', 'Streaming_Videos', 'Churn')

# Convert to factors for training data
data1.train[factor_columns] <- lapply(data1.train[factor_columns], as.factor)

# Convert to factors for test data
X.test[factor_columns] <- lapply(X.test[factor_columns], as.factor)

# Check that the types match
sapply(data1.train[factor_columns], class)
sapply(X.test[factor_columns], class)

```

```{r}

# Check column names in both train and test data
print(names(data1.train))
print(names(X.test))

```

```{r}
# Ensure X.test contains only the predictors (exclude 'Monthly_Charges')
X.test <- data1.test[, -which(names(data1.test) == "Monthly_Charges")]
```

## Fitting The Model

We fit Support Vector Regression (SVR) models with both linear and radial basis function (RBF) kernels, as well as an ordinary least squares (OLS) regression model (with step-wise variable selection). We will the evaluate the performance of these three regression models using mean squared error (MSE) and mean absolute error (MAE)

The residual plot (top left panel) shows no curving which indicate that the model perform well.However, even if the model performs well, will look at the transformed log Monthly_Charges and see if we get better fit.

```{r}

# Step 1: Train the model
lse.fit <- lm(Monthly_Charges ~ ., data = data1.train)  # Initial model
AIC.fit <- stepAIC(lse.fit, direction = "both", trace = FALSE)  # Stepwise AIC model

# Step 2: Make predictions on the test set (Ensure X.test is the right structure)
X.test <- data1.test[, -which(names(data1.test) == "Monthly_Charges")]  # Exclude target
pred.lse <- predict(AIC.fit, X.test)

# Step 3: Evaluate the model
mse.lse <- mean((y.test - pred.lse)^2)    # Mean Square Error
mae.lse <- mean(abs(y.test - pred.lse))   # Mean Absolute Error

par(mfrow=c(2,2), mar=c(2,2,2,2))
plot(AIC.fit)

```

```{r}
#str(data1.train)  # Inspect the structure of the dataset
```



# Fitting The log Transformed Model 

No obvious difference can be seen from the log transfer model which indicate that the log transformation is ineffective.

```{r}

data1.train[] <- lapply(data1.train, function(x) if (is.character(x)) as.factor(x) else x)
X.test[] <- lapply(X.test, function(x) if (is.character(x)) as.factor(x) else x)

data1.train[] <- lapply(data1.train, function(x) if (is.factor(x)) as.numeric(x) else x)
X.test[] <- lapply(X.test, function(x) if (is.factor(x)) as.numeric(x) else x)

#str(data1.train)  # Inspect the structure of the dataset
#str(X.test)  # Inspect the structure of the dataset
```

  

```{r}

data1.train$Monthly_Charges_log <- log(data1.train$Monthly_Charges + 1)  

X.test <- data1.test[, -which(names(data1.test) == "Monthly_Charges")]  # Exclude target

# Make predictions
pred.lse <- predict(AIC.fit, X.test)

# Assign the predicted values to the X.test dataset as the new Monthly_Charges
X.test$Monthly_Charges <- pred.lse

# Apply log transformation with safeguard for zero/negative values
X.test$Monthly_Charges_log <- log(pmax(X.test$Monthly_Charges, 0) + 1)  

# Evaluate the model
mse.lse <- mean((y.test - pred.lse)^2)    # Mean Square Error
mae.lse <- mean(abs(y.test - pred.lse))   # Mean Absolute Error

# Plot the diagnostics
par(mfrow = c(2, 2), mar = c(2, 2, 2, 2))
plot(AIC.fit)  # Plot diagnostics for the model

```

  + comparing the performance of different models, table below, shows that Linear.SVR is the best performing model with the lowest mean squared (MSE) error and Mean Absolute Error (MAE)
  
  
```{r}

Performance <- data.frame(RBF.SVR=c(mse.RBF, mae.RBF),
                          Linear.SVR = c(mse.lin, mae.lin),
                          LSE.Reg =c(mse.lse, mae.lse),
                          LSE.Reg.log = c(mse.lse, mae.lse))
row.names(Performance) <- c("MSE", "MAE")
##
pander(Performance)

```

